\documentclass[assignment]{tmanotes}
\settmacourse{TMA4220}
{Numerical Solution of Partial Differential Equations Using Finite Element Methods}
{Numerical Solution of PDEs Using Finite Element Methods}

\input{../macros}

\bibliographystyle{plain}
\title{Project II -- Applications in three dimensions}
\date{}

\begin{document}
\maketitle

The deadline for this assignment is 25. November 2018 and counts towards 20\% of the final grade.
The delivery consists of a short report describing the method and presenting the results, together with a source code in MATLAB or Python.
Working in pairs is possible but not compulsory.

\medskip
A short presentation of the results will be organized on 20. November 2018 to share your results and as a farewell party.

\section{Projects}

The proposed topics are build on the work done for the Helmoltz equation for Project I.

\medskip
Any of the traditional tracks can be chosen, and one additional:
\begin{itemize}
\item Making a princess cake: solving the Heat Equation, Section 5.1 \cite{P2}.
\item Structural analysis, Section 5.2 \cite{P2}.
\item Vibration analysis, Section 5.3 \cite{P2}.
\item Poisson with varying coefficient, Section 4 \cite{P1} plus extension.
\end{itemize}

\medskip
The last option alternative track is to implement the Poisson Equation in three dimensions with varying coefficient, by first extending the implementation in the constant coefficient case Section 4. \cite{P1} to three dimensions, then choosing a Finite Element discretization for an analytic solution with varying coefficient.
Depending on your interest or Master Thesis work we can discuss which Finite Element discretization to use.

\medskip
For integration rules in three dimensions, refer to \cite{P1}.

\bibliography{../main}

\end{document}

